The Iterative Signature Algorithm

### 2009-05-29 isa2/inst/doc/ISA_tutorial.Rnw
isa2: updated vignettes.

Fig.~\ref{fig:twoclusters} shows possibly the simplest
example of a (rather artificial) data matrix with very strong modular
structure. It is a $20\times 20$ matrix and has two correlated blocks,
each of them of size $10\times 10$.
\begin{figure}
<<toymatrix,echo=FALSE,fig=TRUE,width=6,height=3>>=
library(isa2)
library(Matrix)
data <- data2 <- isa.in.silico(20,20,2,10,10)[[1]]
perm1 <- sample(seq_len(nrow(data)))
data2 <- data2[perm1,]
perm2 <- sample(seq_len(ncol(data)))
data2 <- data2[,perm2]

@@

This iteration is performed until it converges, i.e. $r_{i-1}$
and $r_i$ are \emph{close}, and $c_{i-1}$ and
$c_i$ are also close. The convergence criteria,
i.e. what \emph{close} means is by default defined by high Pearson
correlation.

It is very possible that the ISA finds the same modules more than once;
two or more seeds might converge to the same module. The function
\Rfunction{isa.unique} eliminates every module from the result of 
\Rfunction{isa.iterate} that is very similar (in terms of
Pearson correlation) to the one that was already found before.

@@

The two main parameters of ISA are the two thresholds (one for the
rows and one for the columns). They basically define the stringency of
the modules. If the row threshold is high, then the modules will have
very similar rows. If it is mild, then modules will be bigger, with
less similar rows than in the first case. The same applied to the
column threshold and the columns of the modules.

@@

It is possible to use non-random seeds as well, if you have some
knowledge about the data or are interested in a particular subset of
rows/columns, then you can feed in your seeds into the
\Rfunction{isa.iterate} function directly. In this case the
algorithm is deterministic, for the same seed you will always get the
same results. This can be considered as a semi-supervised approach.

@@

On in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{isa.normalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$ and
centering and scaling its rows (see \Rfunction{scale}). $E_c$ is
calculated by centering and scaling the rows of $E$. $E_r$ is
used to calculate the column signature of rows and $E_c$ is used
to calculate the signature of the columns.

@@

\subsection{Robustness}
As ISA is an unsupervised algorithm, it may very well find some
modules, even if you feed in noise as an input matrix. To avoid these
spurious modules we defined a robustness measure, a single number for
a module that measures how well the rows and the columns are
correlated. The robustness score is a generalization of the first
singular value of the matrix. If there would be no thresholding, then
the ISA would (almost always) converge to the leading SVD vector and
the robustness score would be the corresponding singular value.

It recommended that the user uses \Rfunction{isa.filter.robust} to
run ISA on the scrambled input matrix with the same threshold
parameters and then drop every module, which has a robustness score
lower than the highest robustness score among modules found in the
scrambled data.

@@

The simplest way to use ISA on your data is by calling the
\Rfunction{isa} function with your input matrix as the single
arguments. This function uses random seeding and has three optional
arguments: 
\begin{description}
  \item[\Rargument{thr.row}] A numeric vector, the row thresholds to
    use. By default this is $1,1.5,2,2.5,3$. 
  \item[\Rargument{thr.col}] A numeric vector, the column thresholds
    to use. By default this is $1,1.5,2,2.5,3$.
  \item[\Rargument{no.seeds}] The number of random seed vectors to
    generate. By default 100 seeds are generated.
\end{description}

@@

Let us see an example. We load the \Rpackage{isa2} generate some in
silico data with the \Rfunction{isa.in.silico} function first:
<<insilico,cache=TRUE>>=
data <- isa.in.silico(200,200,2,50,50)[[1]]
@
The function call above generates the data matrix in
Fig.~\ref{fig:twoclusters}. The first two arguments define the size of
the data matrix ($200\times 200$), the third the number of modules (two)
and the fourth and fifth give the size of the modules, both are
$50\times 50$ submatrices in this case. By default
\Rfunction{isa.in.silico} creates non-overlapping modules with some
background noise.

\Rfunction{isa.in.silico} returns a list with a list with three
elements, we are most interested in the first now, that is the actual
artificial data matrix. The second and the entries in the list give
the correct module memberships of the elements, see the manual page of
\Rfunction{isa.in.silico} for the details. Let's take a look at the
input data, see Fig.~\ref{fig:block}.

@@

The \Rfunction{isa} function returns a list with four
elements. \listentry{rows} and \listentry{columns} define the
modules. The non-zero elements in the first column of both define the
first module, the second columns define the second module, etc.
The number of columns in \listentry{rows} (and \listentry{columns})
correspond to the number of modules.
<<no.mods.>>=
ncol(modules$rows)
@
ISA has found \Sexpr{ncol(modules$rows)} %$
modules in the artificial data. This seem a bit surprising, as we
expected only two modules, so let us take a look at the modules. We
simply plot the row scores of the individual modules
(Fig.~\ref{fig:simpleplot}). The first two modules corresponds to the
two blocks in the input matrix. The third module is the union of them,
with opposite score signs. The \Rfunction{isa} function finds modules
containing anticorrelated rows/columns, as long as the
anti-correlation is coordinated. I.e. in the third module, the first
50 rows \emph{always} behave in an opposite way compared to the second
50 rows. This is why they are collected into a single module.

@@

The \listentry{rundata} entry of the return value of \Rfunction{isa}
contains data about the ISA run, all the parameters (the default
values in our case) are included here. It has the following members:
<<isaresult.rundata>>=
names(modules$rundata)
@ 
See the documentation of \Rfunction{isa} for more about these.

Finally, the listentry{seeddata} entry is a data frame, it contains
various data about the individual modules or the seeds that were used
to find the module. There is one row for each module. It has the
following columns:
<<seedcols>>=
colnames(modules$seeddata)
@ 
The columns \listentry{thr.row} and \listentry{thr.col} contain the
row and columns thresholds that were used to find the module, while
the \listentry{rob} column contains its robustness score:
<<seeddata>>=
modules$seeddata
@

@@

In this section we will perform each steps of the ISA analysis
individually. This makes sense only if the user wants to adjust the
parameters of some steps.

@@

To run the ISA, we need to create the starting seeds first. The
\Rfunction{isa} function uses sparse random seeds, generated by
\Rfunction{generate.seeds}. We also have the choice of using
non-random seeds, e.g. if you have a matrix with gene expression data
measured on many samples you can use ``gene'' seeds that correspond to
biological pathways; in this case ISA does a biased search to find
modules related to the input pathways. Or, in a case/control
experiment one can use seed vectors that correspond to cases and look
for modules that are differentially regulated in the control and the
case samples.

@@

The \Rfunction{isa.unique} function finds duplicate or very similar
modules in a list returned by \Rfunction{isa.iterate} and keeps only a
single one of them. It also eliminates all-zero modules and seeds
that did not converge, the latter result all \verb+NA+
columns in the \listentry{rows} and \listentry{columns} matrix:
<<unique>>=
modules2 <- isa.unique(normed.data, modules, cor.limit=0.9)
@
The \Rargument{cor.limit} argument specifies the correlation limit
above which two modules are considered to be the same. Let us see how
many modules are left:
<<isa.unique.res>>=
ncol(modules2$rows)
@

@@

One can use the robustness measure to filter out list of modules, in
the following way. We permute the input matrix and run the same
module finding procedure that we did for the original matrix. Then we
calculate the robustness scores for both the real and the scrambled
modules and eliminate the ``real'' modules having a robustness score
that is less than the score of at least one scrambled module. Note,
that this filtering can be done only if both sets of modules were
found using the same threshold parameters. The procedure is
implemented in the \Rfunction{isa.filter.robust} function:
<<filter.robust>>=
modules3 <- isa.filter.robust(data[[1]], normed.data, modules2, 
                              perms=1, row.seeds=row.seeds)
@
We use the very same row seeds, as for the original matrix; this is
not strictly requires. The \Rargument{perms} argument sets the number
of permutations to perform.

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=7,height=4>>=
<<robbox3>>
@
\caption{The robustness scores of ISA modules after filtering out some
  modules with low robustness. It is still obvious that ten modules
  have a significantly higher score than the rest.}
\label{fig:robbox2}
\end{figure}

The scatterplot shows that ten modules have in fact higher
robustness scores than the rest. Let us check that these are indeed the
ten ``real'' modules of the data.
<<bestmods>>=
bestmods <- which(modules3$seeddata$rob > 27.5)
mod.cor <- pmin(cor(modules3$rows[,bestmods]!=0, data[[2]]),
                cor(modules3$columns[,bestmods]!=0, data[[3]]))
apply(mod.cor, 1, max)
apply(mod.cor, 2, max)
@

Indeed, we have found the ten real modules, the ISA is both specific
and sensitive. Of course for the analysis of real data it is much more
difficult to filter the modules found by the biclustering algorithm,
but the robustness based filtering still helps getting rid of spurious
modules.

Finally, let us take a look at some of the other modules, that were
also kept after the robustness-based filtering. Here we plot nine of
them.
<<othermods.pre,echo=FALSE,results=hide>>=
png("othermods.png", 400, 400)
@ 
<<othermods>>=
othermods <- which(modules3$seeddata$rob < 27.5)[1:5]
print(plotModules(modules3, othermods, data=data[[1]]))
@ 
<<othermods.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{othermods}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  in fact two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

@@

Visualizing overlapping biclusters is quite a challenging task. We
show simple methods that usually visualize a single bicluster 
at a time. For some of these we will use the \Rpackage{biclust} R
package.

@@

There are some examples in this document that show how to create image
plots for the modules and potentially also the input data. See
e.g. Figs~\ref{fig:othermods} and
\ref{fig:overlapplot}. These plots use the \Rfunction{plotModules}
function, that directly takes the output of \Rfunction{isa} (and other
functions with the same output format: \Rfunction{isa.iterate},
Rfunction{isa.unique}, etc.

In fact \Rfunction{plotModules} calls the
\Rfunction{images} function to do its job. \Rfunction{images} takes a
list of matrices, see Fig.~\ref{fig:noise} for an example.

@@

The \Rfunction{parallelCoordinates} function of the \Rpackage{biclust}
package plots the profile of the rows (or columns) that are included
in a bicluster with a different color. These plots visualize the
difference between bicluster and the rest of the matrix, see
Fig.\ref{fig:parallel} for an example.
<<parallel.pre,include=TRUE>>=
parallelCoordinates(five.mods[[1]], Bc, number=1, plotBoth=TRUE)
@ 
\begin{figure}
\centering
\setkeys{Gin}{width=0.6\textwidth}
<<parallel,fig=TRUE,echo=FALSE,results=hide>>=
<<parallel.pre>>
@ 
\caption{Profile plots for the first bicluster found by ISA. In this
  artificial data set there is a clear distinction between the
  background and the rows/columns of the module.}
\label{fig:parallel}
\end{figure}

@@

The \Rfunction{plotclust} function of the \Rpackage{biclust} package
creates barplots for one or more biclusters. A single bar is the
mean of the columns of the bicluster for a given row of the bicluster;
or the mean of the columns of the background (i.e. the rest of the
input matrix). The bigger the difference between the bars of the two
colors, the better the bicluster. The results of the following code
are in Fig.~\ref{fig:plotclust}.
<<plotclust.pre,include=FALSE>>=
plotclust(Bc, five.mods[[1]], Titel="")
@
\begin{figure}
\centering
\setkeys{Gin}{width=0.6\textwidth}
<<plotclust,fig=TRUE,echo=FALSE,results=hide>>=
<<plotclust.pre>>
@ 
\caption{Bar plots to show the difference between the biclusters and
  the rest of the input matrix, for the first five modules found by
  ISA.}
\label{fig:plotclust}
\end{figure}

@@

Let's create boxplots for the different noise levels.

<<sensbox,include=FALSE,results=hide>>=
boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<sensbox>>
@ 
\caption{The sensitivity of the ISA in the function of noise. The
  sensitivity score is calculated by finding the best (in terms of
  Pearson correlation) ISA module for each block in the input matrix.
  ISA is very sensitive to relatively large noise levels.
}
\label{fig:sensbox}
\end{figure}

To test the specificity, we create boxplots for the correlation
coefficients of all modules found by ISA and their closest real
module.
<<spec>>=
spec <- lapply(seq_along(modules), function(i) {
  cc <- pmin(cor(modules[[i]]$rows != 0, data[[i]][[2]]),
             cor(modules[[i]]$columns != 0, data[[i]][[3]]))
  apply(cc, 1, max)
})
@ 
<<specbox,include=FALSE,results=hide>>=
boxplot(spec, names=noise, xlab="Noise level", ylab="Specificity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<specbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<specbox>>
@
\caption{Specificity of the ISA modules, in the function of
  noise. Specificity was calculated by finding the best -- again,
  by highest Pearson correlation -- matching block in the input matrix
  for each ISA module. These correlation coefficients are plotted in a
  boxplot for each noise level.
  Specificity of ISA decreases approximately linearly with the
  increase of noise.
}
\label{fig:specbox}
\end{figure}

@@

As expected the ISA modules are in general bigger with the mild
thresholds. ISA correctly identifies the two overlapping modules and
also the union of them. With the stricter thresholds it find the
non-overlapping parts of the two modules, plus their overlap as a
separate module.

>>>>

Fig.~\ref{fig:twoclusters} shows possibly the simplest
example of a (rather artificial) data matrix with very strong modular
structure. It is a $20\times 20$ matrix and (after reordering its rows
and columns) it has two correlated blocks,
each of them of size $10\times 10$.
\begin{figure}
<<toymatrix,echo=FALSE,fig=TRUE,width=6,height=3>>=
library(isa2)
library(Matrix)
data <- data2 <- isa.in.silico(20,20,2,10,10)[[1]]
perm1 <- sample(seq_len(nrow(data)))
data2 <- data2[perm1,]
perm2 <- sample(seq_len(ncol(data)))
data2 <- data2[,perm2]

@@

This iteration is performed until it converges, i.e. $r_{i-1}$
and $r_i$ are \emph{close}, and $c_{i-1}$ and
$c_i$ are also close. The convergence criteria,
i.e. what \emph{close} means, is by default defined by high Pearson
correlation.

It is very possible that the ISA finds the same module more than once;
two or more seeds might converge to the same module. The function
\Rfunction{isa.unique} eliminates every module from the result of 
\Rfunction{isa.iterate} that is very similar (in terms of
Pearson correlation) to the one that was already found before.

@@

The two main parameters of ISA are the two thresholds (one for the
rows and one for the columns). They basically define the stringency of
the modules. If the row threshold is high, then the modules will have
very similar rows. If it is mild, then modules will be bigger, with
less similar rows than in the first case. The same applies to the
column threshold and the columns of the modules.

@@

It is possible to use non-random seeds as well. If you have some
knowledge about the data or are interested in a particular subset of
rows/columns, then you can feed in your seeds into the
\Rfunction{isa.iterate} function directly. In this case the
algorithm is deterministic, for the same seed you will always get the
same results. Using smart (i.e. non-random) seeds can be considered as
a semi-supervised approach.

@@

Using in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{isa.normalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$ and
centering and scaling its rows (see the \Rfunction{scale} R function). $E_c$ is
calculated by centering and scaling the rows of $E$. $E_r$ is
used to calculate the column signature of rows and $E_c$ is used
to calculate the signature of the columns.

@@

\subsection{Robustness}
As ISA is an unsupervised algorithm, it may very well find some
modules, even if you feed in noise as an input matrix. To avoid these
spurious modules we defined a robustness measure, a single number for
a module that measures how well the rows and the columns are
correlated. The robustness score is a generalization of the first
singular value of the matrix. If there would be no thresholding during
the ISA iteration, then
the ISA would (almost always) converge to the leading SVD vector and
the robustness score would be the corresponding singular value.

It is recommended that the user uses \Rfunction{isa.filter.robust} to
run ISA on the scrambled input matrix with the same threshold
parameters and then drop every module, which has a robustness score
lower than the highest robustness score among modules found in the
scrambled data.

@@

The simplest way to use ISA on your data is by calling the
\Rfunction{isa} function with your input matrix as the single
argument. This function uses random seeding and has three optional
arguments: 
\begin{description}
  \item[\Rargument{thr.row}] A numeric vector, the row thresholds to
    use. By default this is
    \Sexpr{paste(eval(formals(isa2:::isa.default)$thr.row), collapse=", ")}. % $
  \item[\Rargument{thr.col}] A numeric vector, the column thresholds
    to use. By default this is
    \Sexpr{paste(eval(formals(isa2:::isa.default)$thr.col), collapse=", ")}. % $
  \item[\Rargument{no.seeds}] The number of random seed vectors to
    generate. By default 
    \Sexpr{formals(isa2:::isa.default)$no.seeds} seeds are generated. % $
\end{description}

@@

Let us see an example. We load the \Rpackage{isa2} package, and
generate some in 
silico data with the \Rfunction{isa.in.silico} function first:
<<insilico,cache=TRUE>>=
library(isa2)
data <- isa.in.silico(200,200,2,50,50)[[1]]
@
The function call above generates a data matrix similar to the one in
Fig.~\ref{fig:twoclusters}, but a bigger one. The first two arguments define the size of
the data matrix ($200\times 200$), the third the number of modules (two)
and the fourth and fifth give the size of the modules, both are
$50\times 50$ submatrices in this case. By default
\Rfunction{isa.in.silico} creates non-overlapping modules with some
background noise.

\Rfunction{isa.in.silico} returns a list with three
elements, we are most interested in the first now, that is the actual
artificial data matrix. The second and third entries in the list give
the correct module memberships of the elements, see the manual page of
\Rfunction{isa.in.silico} for the details. Let's take a look at the
input data, see Fig.~\ref{fig:block}.

@@

The \Rfunction{isa} function returns a list with four
elements. \listentry{rows} and \listentry{columns} define the
modules. The non-zero elements in the first column of both define the
first module, the second columns define the second module, etc.
The number of columns in \listentry{rows} (and \listentry{columns})
correspond to the number of modules.
<<no.mods.>>=
ncol(modules$rows)
@
ISA has found \Sexpr{ncol(modules$rows)} %$
modules in the artificial data. This seem a bit surprising, as we
expected only two modules, so let us take a look at the modules. We
simply plot the row scores of the individual modules
(Fig.~\ref{fig:simpleplot}). The first two modules corresponds to the
two blocks in the input matrix. The third module is the union of them,
with opposite score signs. The \Rfunction{isa} function also finds modules
containing anticorrelated rows/columns, as long as the
anti-correlation is coordinated. I.e. in the third module, the first
50 rows \emph{always} behave in an opposite way compared to the second
50 rows. This is why they are collected into a single module. The
following code creates Fig.~\ref{fig:simpleplot}.

@@

The \listentry{rundata} entry of the return value of \Rfunction{isa}
contains information about the ISA run, all the parameters (the default
values in our case) are included here. It has the following members:
<<isaresult.rundata>>=
names(modules$rundata)
@ 
See the documentation of \Rfunction{isa} for more about these.

Finally, the \listentry{seeddata} entry is a data frame, it contains
various information about the individual modules and the seeds that
were used to find them. There is one row for each module. It has the
following columns:
<<seedcols>>=
colnames(modules$seeddata)
@ 
The columns \listentry{thr.row} and \listentry{thr.col} contain the
row and columns thresholds that were used to find the module, while
the \listentry{rob} column contains its robustness score:
<<seeddata>>=
modules$seeddata
@
See the documentation of the \Rfunction{isa} function for more
information about these and other fields.

@@

In this section we will perform each steps of the ISA analysis
individually. This makes sense only if the user wants to adjust the
parameters of some steps. Otherwise a simple call to the
\Rfunction{isa} function would do.

@@

To run the ISA, we need to create the starting seeds first. The
\Rfunction{isa} function uses sparse random seeds, generated by
\Rfunction{generate.seeds}. We also have the choice of using
non-random seeds, e.g. if you have a matrix with gene expression data
measured on many samples you can use ``gene'' seeds that correspond to
biological pathways; in this case ISA does a biased search to find
modules related to the input pathways. Or, in a case/control
experiment one can use seed vectors that correspond to cases and look
for modules that are different in the control and the
case samples.

@@

The \Rfunction{isa.unique} function finds duplicate or very similar
modules in a list returned by \Rfunction{isa.iterate} and keeps only a
single one of them. It also eliminates all-zero modules and seeds
that did not converge, \verb+NA+
columns in the \listentry{rows} and \listentry{columns} matrix
correspond to non-convegent seeds:
<<unique>>=
modules2 <- isa.unique(normed.data, modules, cor.limit=0.9)
@
The \Rargument{cor.limit} argument specifies the correlation limit
above which two modules are considered to be the same. Let us see how
many modules are left:
<<isa.unique.res>>=
ncol(modules2$rows)
@

@@

One can use the robustness measure to filter out list of modules, in
the following way. We permute the input matrix and run the same
module finding procedure that we did for the original matrix. Then we
calculate the robustness scores for both the real and the scrambled
modules and eliminate the ``real'' modules having a robustness score
that is less than the score of at least one scrambled module. Note,
that this filtering can be done only if both sets of modules were
found using the same threshold parameters. The procedure is
implemented in the \Rfunction{isa.filter.robust} function:
<<filter.robust>>=
modules3 <- isa.filter.robust(data[[1]], normed.data, modules2, 
                              perms=1, row.seeds=row.seeds)
@
We used the very same row seeds, as for the original matrix; this is
not strictly required. The \Rargument{perms} argument sets the number
of permutations to perform.

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=7,height=4>>=
<<robbox3>>
@
\caption{The robustness scores of ISA modules after filtering out some
  modules with low robustness. Ten modules
  have a higher score than the rest.}
\label{fig:robbox2}
\end{figure}

The scatterplot shows that ten modules have in fact higher
robustness scores than the rest. Let us check that these are indeed the
ten ``real'' modules of the data.
<<bestmods>>=
bestmods <- order(modules3$seeddata$rob, decreasing=TRUE)[1:10]
mod.cor <- pmin(cor(modules3$rows[,bestmods]!=0, data[[2]]),
                cor(modules3$columns[,bestmods]!=0, data[[3]]))
apply(mod.cor, 1, max)
apply(mod.cor, 2, max)
@

Indeed, we have found the ten real modules, the ISA is both specific
and sensitive. If one analyzes real data, then it is much more
difficult to filter the modules found by the biclustering algorithm,
but the robustness based filtering still helps getting rid of spurious
modules.

Finally, let us take a look at some of the other modules, that were
also kept after the robustness-based filtering. Here we plot five of
them, together with the input matrix.
<<othermods.pre,echo=FALSE,results=hide>>=
png("othermods.png", 400, 400)
@ 
<<othermods>>=
othermods <- order(modules3$seeddata$rob)[1:5]
print(plotModules(modules3, othermods, data=data[[1]]))
@ 
<<othermods.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{othermods}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

@@

Visualizing overlapping biclusters is a challenging task. We
show simple methods that usually visualize a single bicluster 
at a time. For some of these we will use the \Rpackage{biclust} R
package.

@@

There are some examples in this document that show how to create image
plots for the modules and potentially also the input data. See
e.g. Figs~\ref{fig:othermods} and
\ref{fig:overlapplot}. These plots use the \Rfunction{plotModules}
function, that directly takes the output of \Rfunction{isa} (and other
functions with the same output format: \Rfunction{isa.iterate},
\Rfunction{isa.unique}, etc. (In fact \Rfunction{plotModules} calls the
\Rfunction{images} function to do its job. \Rfunction{images} takes a
list of matrices, see Fig.~\ref{fig:noise} for an example.)

@@

The \Rfunction{parallelCoordinates} function of the \Rpackage{biclust}
package plots the profile of the rows (or columns) that are included
in a bicluster with a different color. These plots visualize the
difference between the bicluster and the rest of the matrix, see
Fig.~\ref{fig:parallel} for an example.
<<parallel.pre,include=TRUE>>=
parallelCoordinates(five.mods[[1]], Bc, number=1, plotBoth=TRUE)
@ 
\begin{figure}
\centering
\setkeys{Gin}{width=0.6\textwidth}
<<parallel,fig=TRUE,echo=FALSE,results=hide>>=
<<parallel.pre>>
@ 
\caption{Profile plots for the first bicluster found by ISA. In this
  artificial data set there is a clear distinction between the
  background and the rows/columns of the module.}
\label{fig:parallel}
\end{figure}

@@

The \Rfunction{plotclust} function of the \Rpackage{biclust} package
creates barplots for one or more biclusters. A single bar is the
mean of the columns of the bicluster for a given row of the bicluster;
or the mean of the columns of the background (i.e. the rest of the
input matrix). The bigger the difference between the bars of the two
colors, the better the bicluster. The results of the following code
are in Fig.~\ref{fig:plotclust}.
<<plotclust.pre,include=FALSE>>=
plotclust(Bc, five.mods[[1]], Titel="")
@
\begin{figure}
\centering
\setkeys{Gin}{width=0.8\textwidth}
<<plotclust,fig=TRUE,echo=FALSE,results=hide,width=10,height=7>>=
<<plotclust.pre>>
@ 
\caption{Bar plots show the difference between the biclusters and
  the rest of the input matrix, for the first five modules found by
  ISA.}
\label{fig:plotclust}
\end{figure}

@@

Let's create boxplots for the different noise
levels. (Fig.~\ref{fig:sensbox}.)

<<sensbox,include=FALSE,results=hide>>=
boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<sensbox>>
@ 
\caption{The sensitivity of the ISA in the function of noise. The
  sensitivity score is calculated by finding the best (in terms of
  Pearson correlation) ISA module for each block in the input matrix.
  ISA performs very well, even in the presence of relatively high
  noise levels.
}
\label{fig:sensbox}
\end{figure}

To test the specificity, we create boxplots for the correlation
coefficients of all modules found by ISA and their closest real
module. (Fig.~\ref{fig:specbox}.)
<<spec>>=
spec <- lapply(seq_along(modules), function(i) {
  cc <- pmin(cor(modules[[i]]$rows != 0, data[[i]][[2]]),
             cor(modules[[i]]$columns != 0, data[[i]][[3]]))
  apply(cc, 1, max)
})
@ 
<<specbox,include=FALSE,results=hide>>=
boxplot(spec, names=noise, xlab="Noise level", ylab="Specificity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<specbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<specbox>>
@
\caption{Specificity of the ISA modules, in the function of
  noise. Specificity was calculated by finding the best -- again,
  by highest Pearson correlation -- matching block in the input matrix
  for each ISA module. These correlation coefficients are plotted in a
  boxplot for each noise level.
  Specificity of ISA decreases approximately linearly with the
  increase of noise.
}
\label{fig:specbox}
\end{figure}

@@

As expected the ISA modules are in general bigger with the mild
thresholds. ISA correctly identifies the two overlapping modules and
also the union of them. With the stricter thresholds it finds the
non-overlapping parts of the two modules, plus their overlap as a
separate module.
